Can dike grassland restoration profit from ideal seed mixture-substrate combinations?

### 2023-02-18 manuscript.Rmd
Update manuscript.Rmd

Under the substrate, a 5 cm thick drainage layer of gravel (0-16 mm) was installed. Soil samples of the three substrates were tested on both expositions by mixing several sub-samples from different plots. The sand admixture changed the soil texture, increased the C/N ratio, reduced Calciumcarbonat, but did hardly changed the quite high pH values (Table \@ref(tab:substrate)). Phosphate and potassium were rather scarce for agricultural soils and Magnesium rather high [@lfl2022]. The vegetation was surveyed in June or July 2018--2021 [@BraunBlanquet.1964] and Londo scale was used [@londo1976]. The establishment rates of species are recorded in Appendix A3. `r word_spec("(Add mean values, range, how many of species pool established, etc.)")` <!--# add more infos -->

@@

The ordination presents the experimental plots also in context with reference sites. The reference sites have a bigger variation than the seed mixtures and are close to the seed mixtures but hardly overlap. The positions of reference sites are shifted to the left in comparison to the seed mixtures which means in the direction of early successional states. Though, they are still different from the negative references of ruderal, dry, anthropogenic vegetation. Negative references were only available on the south exposition and they are located in the NMDS between the positive reference sites and the state of restored plots in 2021. Hay meadow-seed mixtures led to a closer development to hay meadow references than dry grasslands to their references (Figure \@ref(fig:effects), \@ref(fig:recovery)) this is especially the case on the north exposition (Figure \@ref(fig:nmds)).

@@

The vegetation of the dike experiment developed in the right direction, but with reduced velocity and significantly and lasting worse development of the south exposition. Sand admixture of 25% was sufficient to reach a positive effect. All in all, it was hardly possible to identify an ideal seed mixture-substrate combination. Though, hay meadows developed closer to its reference on the north exposition than dry grasslands.

@@

As expected, sand admixture led to higher persistence of sown species but only on the north exposition and the effect size of about 6% in the fourth year after restoration is rather small. In addition, 50% admixture did not further improve the result. Restoration on agricultural soils can have limited success due to high nutrient loads [@walker2004] but can lead to positive results, if mixed with mineral material as sand and gravel [@chenot-lescure2022]. Substrate depth did not significantly affect persistence or Favorable Conservation Status (FCS) similarly to earlier studies [@baer2004]. Larger differences of soil depth might be necessary to create negative effects by greater soil depths as was shown for prairies [@dornbush2010]. Neither, seed density had an effect on persistence and FCS which fits to result of @Kaulfuss.2022, who found out that a certain amount of seeds are necessary but higher densities do not further improve establishment of target species but have rather a slight negative effect.

The vegetation on the south exposition performed consistently and significantly worse than the north exposition and had a different species composition which was also shown by @Batori.2016 in Hungary. Here, the differences might have some unwanted experimental reasons: the geotextile, which had been implemented on the south exposition, was removed after two weeks because of the death of several protected lizards. This led to the drying of many seedlings why the distinction of the seed mixture types might be leveled on the south exposition. Moreover, the intense drought of 2018 and 2019 [@hari2020] might have decreased germination and establishment of sown species because a moist spring with a mesic temperature improves establishment [@Larson.2021; @orrock2023]. The interesting point here is the lasting negative effect which suggests a legacy effect by unfavorable weather conditions after sowing as it was observed by other studies [@Stuble.2017; @Groves.2020]. The problems during the establishment phase might have led to an alternative trajectory and could probably only be met by resowing.

@@

Our aim was to identify a ideal seed mixture-substrate combination in respect of restoration effectiveness and biodiversity. For evaluating effectiveness, we measured the establishment and persistence of the sown species and for fostering plant biodiversity in general, we measured the FCS. Though, we could not identify an interaction effect for neither of the indices. We would have expected a better performance of hay-meadow seed mixtures with lower sand admixture and for dry grasslands with higher sand admixture. Our results suggest that, at least after four years, the substrate conditions are within the range of both seed mixture types (hay meadows, dry grasslands). Although, both types are clearly plant-sociologically and functionally distinct, they are still relative closely related why they contain shared species and could develop on similar site conditions [Figure @ref(fig:traits); @Oberdorfer.1993; @Oberdorfer.1993b]. Other grassland studies could identify more or less clear interactions of opposing habitat preferences or functional traits along the gradients of productivity, moisture, and nutrients [@Zirbel.2020; @Freitag.2021; @Kaulfuss.2022]. However, these two studies did not work with an experimental set up of different seed mixture-substrate combinations, but analyzed the result of habitat and biotic filtering after five and about 15 years, respectively. Furthermore, the non-existence of an ideal combination could be explained by priority effects which means that the species of the sown species could establish earlier than others and preempted the available niches [@Fukami.2015].

@@

The seed mixtures and positive references were similar but hardly overlapped (Figure \@ref(fig:nmds)). The position on the NMDS suggests that the seed mixture represent a latter successional stage than the positive references. The NMDS shows a slightly better adaption of hay meadows to the north exposition than of dry grasslands (Figure @ref(fig:recovery)). This can be expected from the requirements of hay meadows for mesic conditions which can be provided on north exposed dike slopes [@Batori.2020; @Oberdorfer.1993b]. On the south exposition, the plots of hay meadows developed rather to dry grassland references which indicates an inefficient restoration by using an non-adapted seed mixture.

The vegetation developed in general in the right direction but was still distinct from positive references and seed mixtures after four years. It was rather similar to the negative reference of perennial ruderal vegetation. The gap between goal and restoration outcome was also shown for other sowing experiments or restorations [@mitchley2012; @Engst.2016; @Kaulfuss.2022] or for dike vegetation compared with semi-natural references [@Batori.2016]. This result is not surprising since the 'recovery debt' is a general phenomenon of grassland restoration [@MorenoMateos.2017; @jones2018].

@@

Our results suggest that adapted seed mixtures can increase restoration efficiency by sowing hay meadows on the north exposition and not on the south exposition of dikes. Furthermore, to reduce nutrient load by sand admixture has a clear positive effect, although the effect is small. The question remains if money is best invested in sand admixture or in other restoration measures. It seems not to be necessary to increase seed density on dike slopes, which would reduce restoration costs. As well as the use of more deeper soil depth which would otherwise create costs for removal of topsoil. There was no perfect seed mixture-substrate combination, why we conclude that a variation of seed mixtures types and different soils along restoration sections would promote biodiversity better than a single solution [@bauer2023; @Holl.2022]. The strong and lasting negative effect of drought on the south exposition might require a resowing. To close the recovery debt, the adaption of the management might be promising since it is a relevant factor aside of restoration approach and site characteristics [@Grman.2013; @tölgyesi2021]. For example, starting sheep grazing on the restored sites will modify the disturbance regime and improve dispersal. All in all, our results support the finding that dike grasslands can promote biodiversity in agricultural landscapes [@Batori.2020]. Though, the recovery debt highlights that restored grasslands cannot substitute old-growth grasslands [@nerlekar2020; @Buisson.2022].

@@

We would like to thank our project partners Dr. Markus Fischer, Frank Schuster, and Christoph Schwahn (WIGES GmbH) and Klaus Rachl and Stefan Radlmair (Regierung von Niederbayern) for numerous discussions on restoration and management of dike grasslands. Field work was supported by Clemens Berger and Uwe Kleber-Lerchbaumer (Wasserwirtschaftsamt Deggendorf). We thank Holger Paetsch, Simon Reith, Anna Ritter, Jakob Strak, Leonardo H. Teixeira, and Linda Weggler for assisting with the field surveys or soil analyses in 2018--2020. The German Federal Environmental Foundation (DBU) supported MB with a doctoral scholarship. The study has been supported by the TRY initiative on plant traits ([http://www.try-db.org).](http://www.try-db.org).) The TRY initiative and database is hosted, developed and maintained by J. Kattge and G. Boenisch (Max Planck Institute for Biogeochemistry, Jena, Germany). TRY is currently supported by Future Earth/bioDISCOVERY and the German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig.

@@

MB is funded by a doctoral scholarship of the German Federal Environmental Foundation (DBU) (No. 20021/698). The establishment of the experiment and the vegetation surveys were financed by the WIGES GmbH in the years 2018--2020 (No. 80 002 312).

@@

Each plot received an individual set of 20 species with some restrictions to the number of species per functional group. The total species pool for hay meadows was 55 and for dry grassland 58. In total, plots received 16 or 32 g of seeds. All individual seed mixtures are stored in Appendix A3.

```{r tab.id = "seedmix", ft.align = "left", tab.cap = "Seedmix"}
read_csv(here::here("data", "raw", "data_raw_table_1_seedmix.csv"),
         show_col_types = FALSE) %>%
  slice(-1) %>%
  flextable() %>%
  merge_at(part = "header", j = 2:3) %>%
  align(align = "center", part = "all", j = -1) %>%
  add_header_row(top = FALSE, values = c("", "Hay meadow", "Dry grassland", "", "", "")) %>%
  add_header_row(top = FALSE, values = c("", "#", "#", "#", "wt%", "wt%")) %>%
  bold(part = "header", i = 1) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  autofit()
```

@@

Soil samples of the three substrates were analyzed for the fraction \< 2 mm. Soil texture was classified according to the 'Bodenkundliche Kartieranleitung' [@BodenkundlKartieranl.2005]. The pH was measured in CaCl~2~ solution. Plant available phosphorus and potassium were measured in a calcium acetate-lactate extract and magnesium in CaCl~2~ extract. For calculating CaCO~3~, a sub-sample was annealed at 550 °C and the measured C amount multiplied with 8.33. To calculate total N and the C/N ratio, a sub-sample was incinerated at 1000 °C. Lt3 = medium clayey loam; Ls4 = strong sandy loam; Sl3 = medium loamy sand; Sl4 = strong loamy sand.

```{r tab.id = "substrate", ft.align = "left", tab.cap = "Substrate"}
read_csv(here::here("data", "raw", "data_raw_table_3_substrate.csv"),
         show_col_types = FALSE) %>%
  select(-starts_with("lfl")) %>%
  slice(-1) %>%
  flextable() %>%
  bold(part = "header") %>%
  add_header_row(
    top = FALSE,
    values = read_csv(
      here::here("data", "raw", "data_raw_table_3_substrate.csv"),
                 show_col_types = FALSE) %>%
      select(-starts_with("lfl")) %>%
      slice(1)
    ) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  align(align = "center", part = "all", j = -1) %>%
  compose(i = 1, j = 10, part = "header", value = as_paragraph("P", as_sub("2"), "O", as_sub("5"))) %>%
  compose(i = 1, j = 11, part = "header", value = as_paragraph("K", as_sub("2"), "O")) %>%
  compose(i = 1, j = 12, part = "header", value = as_paragraph("Mg", as_sup("2+"))) %>%
  compose(i = 1, j = 14, part = "header", value = as_paragraph("CaCO", as_sub("3"))) %>%
  compose(i = 2, j = 10, part = "header", value = as_paragraph("mg 100 g", as_sup("−1"))) %>%
  compose(i = 2, j = 11, part = "header", value = as_paragraph("mg 100 g", as_sup("−1"))) %>%
  compose(i = 2, j = 12, part = "header", value = as_paragraph("mg 100 g", as_sup("−1"))) %>%
  align(align = "left", part = "footer") %>%
  set_table_properties(layout = "autofit", width = 1)
```

@@

```{r fig.id = "nmds", fig.asp = 0.9696, fig.width = 6.49, fig.cap = "NMDS ordination based on Sørensen dissimilarity of the species composition of 288 plots on dike grasslands over four years after sowing the seed mixtures (black squares).  In addition, 98 positive and negative reference plots were included (filled symbols) from older dike grasslands in the surroundings [@bauer2023] and six plots from sPlotOpen [@Sabatini.2021].  The circles show the standard error of the groups. 2D stress: 0.21."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_6_nmds_800dpi_16.5x16cm.tiff")
  )
```

>>>>

Under the substrate, a 5 cm thick drainage layer of gravel (0-16 mm) was installed. Soil samples of the three substrates were tested on both expositions by mixing several sub-samples from different plots. The sand admixture changed the soil texture, increased the C/N ratio, reduced Calciumcarbonat, but did hardly changed the quite high pH values (Table \@ref(tab:substrate)). Phosphate and potassium were rather scarce for agricultural soils and Magnesium rather high [@lfl2022]. The vegetation was surveyed in June or July 2018--2021 [@BraunBlanquet.1964] and Londo scale was used [@londo1976]. The establishment rates of species are recorded in Appendix A3. In total, 48 species of the species pool of hay meadows (87%) and 46 (79%) of dry grasslands) established by 2021. The species established 31 ± 22% of their sown plot.

@@

The ordination presents the experimental plots also in context with reference sites. The reference sites have a bigger variation than the seed mixtures and are close to the seed mixtures but hardly overlap. The positions of reference sites are shifted to the left in comparison to the seed mixtures which means in the direction of early successional states. Though, they are still different from the negative references of ruderal, dry, anthropogenic vegetation. Negative references were only available on the south exposition and they are located in the NMDS between the positive reference sites and the state of restored plots in 2021. Hay meadow-seed mixtures led to a closer development to hay meadow references than dry grasslands to their references (Figure \@ref(fig:effects), \@ref(fig:recovery)) this is especially the case on the north exposition (Figure \@ref(fig:nmds)). \clearpage \newpage

@@

The vegetation of the dike experiment developed in the desired direction, but with a descelerating recovery rate. On the south exposition, a significant worse effect during establishment phase remained over the complete observation period. All in all, it was hardly possible to identify an ideal seed mixture-substrate combination. Though, hay meadows developed closer to its reference on the north exposition than dry grasslands.

@@

Restoration on agricultural soils can have limited success due to high nutrient loads [@walker2004] but can lead to positive results, if mixed with mineral material as sand and gravel [@chenot-lescure2022]. As expected, sand admixture reduced nutrient loads and led to higher persistence of sown species but 50% admixture did not further increase the effect. In addition, the effect only appeared on the north exposition and the effect size of about 6% in the fourth year after restoration is rather small. Substrate depth did not significantly affect persistence or Favorable Conservation Status (FCS) similarly to earlier studies [@baer2004]. Larger differences in soil depths might be necessary to observe negative effects by thicker substrate layers as it was shown for prairies [@dornbush2010]. Neither, seed density had an effect on persistence and FCS which fits to results of @Kaulfuss.2022, who found out that a certain amount of seeds are necessary for a successful establishment of target species but higher densities do not further improve the result and have rather a slight negative effect.

The vegetation on the south exposition had a different species composition than the north exposition which was also shown by @Batori.2016 in Hungary. Here, the differences might have some unwanted experimental reasons: the geotextile, which had been implemented on the south exposition, was removed after two weeks because of the death of several protected lizards. This led to the drying of many seedlings which was amplified by the intense drought of 2018 and 2019 [@hari2020]. This drought could have decreased germination and establishment of sown species also in general because a moist spring with a mesic temperature improves establishment [@Larson.2021; @orrock2023]. The interesting point here is the lasting negative effect on persistence and FCS which suggests a legacy effect by adverse weather conditions after sowing as it was observed by other studies [@Stuble.2017; @Groves.2020]. The problems during the establishment phase might have led to an alternative trajectory `r word_spec("(reference?)")`<!--# Add ref--> and could probably only be met by resowing. The drying of seedlings might also have leveled the distinction of the seed mixture types on the south exposition.

@@

Our aim was to identify an ideal seed mixture-substrate combination in respect of restoration effectiveness and biodiversity. For evaluating effectiveness, we measured the persistence of the sown species and for fostering plant biodiversity in general, we measured the FCS. Though, we could not identify an interaction effect for neither of the indices. We would have expected a better performance of hay-meadow seed mixtures with lower sand admixture and for dry grasslands with higher sand admixture. Our results suggest that, at least after four years, the substrate conditions are within the range of both seed mixture types (hay meadows, dry grasslands). Although, both types are clearly plant-sociologically and functionally distinct, they are still relative closely related why they contain shared species and could develop on similar site conditions (Figure \@ref(fig:traits)) [@Oberdorfer.1993; @Oberdorfer.1993b].<!--# Check fig ref--> Other grassland studies could identify more or less clear interactions of opposing habitat preferences or functional traits along the gradients of productivity, moisture, and nutrients [@Zirbel.2020; @Freitag.2021; @Kaulfuss.2022]. However, these studies did not work with an experimental set up of different seed mixture-substrate combinations, but analyzed the result of habitat and biotic filtering after one, five and about 15 years, respectively. Furthermore, the non-existence of an ideal combination could be explained by priority effects which means that the species of the non-perfect seed mixture could establish earlier and preempted the available niches for the species of related habitat types [@Fukami.2015].

@@

The seed mixtures and positive reference sites were similar but hardly overlapped (Figure \@ref(fig:nmds)). The position on the NMDS suggests that the seed mixture represent a latter successional stage than the references. The NMDS shows a slightly better adaption of hay meadows to the north exposition than of dry grasslands (Figure \@ref(fig:recovery)). This can be expected from the requirements of hay meadows for mesic conditions which can be provided on north exposed dike slopes [@Batori.2020; @Oberdorfer.1993b]. On the south exposition, the plots of hay meadows developed rather to dry grassland references which indicates an inefficient restoration by using an non-adapted seed mixture.

The vegetation developed in general in the desired direction but was still distinct from positive references and seed mixtures after four years. On the south exposition, the plots were rather similar to the negative reference of dry, ruderal vegetation. The gap between goal and restoration outcome was also shown for other sowing experiments or restorations [@mitchley2012; @Engst.2016; @Kaulfuss.2022] or for dike vegetation compared with semi-natural references [@Batori.2016]. This result is not surprising since the 'recovery debt' is a general phenomenon of grassland restoration [@MorenoMateos.2017; @jones2018].

@@

Our results suggest that adapted seed mixtures can increase restoration efficiency by sowing hay meadows on the north exposition and not on the south exposition of dikes. Furthermore, to reduce nutrient load by sand admixture has a clear positive effect, even if the effect is small. The question which remains is, if sand admixture is the most economical restoration measure to promote diversity on dikes. Seed density seems not to be necessary to be increased on dike slopes and soil depths of 30cm are not adverse compared to 15 cm thick substrates. There was no perfect seed mixture-substrate combination, why we conclude that a variation of seed mixture types and different substrates along restoration sections would promote biodiversity better than a single solution [@bauer2023; @Holl.2022]. The strong and lasting negative effect of drought on the south exposition might require a resowing. To close the recovery debt, the adaption of the management might be promising since it is a crucial factor aside of the restoration approach and the site characteristics [@Grman.2013; @tölgyesi2021]. For example, starting the already existing sheep grazing on the experimental sites will modify the disturbance regime and improve dispersal. All in all, our results support the finding that dike grasslands can promote biodiversity in agricultural landscapes [@Batori.2020]. Though, the stubborn recovery debt highlights that restored grasslands cannot substitute old-growth grasslands [@nerlekar2020; @Buisson.2022].

@@

We would like to thank our project partners Dr. Markus Fischer, Frank Schuster, and Christoph Schwahn (WIGES GmbH) as well as Klaus Rachl and Stefan Radlmair (Lower Bavaria Government) for numerous discussions on restoration and management of dike grasslands. Field work was supported by Clemens Berger and Uwe Kleber-Lerchbaumer (Wasserwirtschaftsamt Deggendorf). We thank Holger Paetsch, Simon Reith, Anna Ritter, Jakob Strak, Leonardo H. Teixeira, and Linda Weggler for assisting with the field surveys or soil analyses in 2018--2020. The German Federal Environmental Foundation (DBU) supported MB with a doctoral scholarship. The study has been supported by the TRY initiative on plant traits ([http://www.try-db.org).](http://www.try-db.org).) The TRY initiative and database is hosted, developed and maintained by J. Kattge and G. Boenisch (Max Planck Institute for Biogeochemistry, Jena, Germany). TRY is currently supported by Future Earth/bioDISCOVERY and the German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig.

@@

MB is funded by a doctoral scholarship of the German Federal Environmental Foundation (DBU) (No. 20021/698). The establishment of the experiment and the vegetation surveys were financed by the WIGES GmbH in the years 2018--2020 (No. 80 002 312). \clearpage \newpage

@@

```{r tab.id = "seedmix", ft.align = "left", tab.cap = "Each plot received an individual set of 20 species with some restrictions to the number of species per functional group. The total species pool for hay meadows was 55 and for dry grassland 58. In total, plots received 16 or 32 g of seeds. All individual seed mixtures are stored in Appendix A3."}
read_csv(here::here("data", "raw", "data_raw_table_1_seedmix.csv"),
         show_col_types = FALSE) %>%
  slice(-1) %>%
  flextable() %>%
  merge_at(part = "header", j = 2:3) %>%
  align(align = "center", part = "all", j = -1) %>%
  add_header_row(top = FALSE, values = c("", "Hay meadow", "Dry grassland", "", "", "")) %>%
  add_header_row(top = FALSE, values = c("", "#", "#", "#", "wt%", "wt%")) %>%
  bold(part = "header", i = 1) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  autofit()
```

@@

Soil samples of the three substrates were analyzed for the fraction \< 2 mm. Soil texture was classified according to the 'Bodenkundliche Kartieranleitung' [@BodenkundlKartieranl.2005]. The pH was measured in CaCl2 solution. Plant available phosphorus and potassium were measured in a calcium acetate-lactate extract and magnesium in CaCl2 extract. For calculating CaCO3, a sub-sample was annealed at 550 °C and the measured C amount multiplied with 8.33. To calculate total N and the C/N ratio, a sub-sample was incinerated at 1000 °C. Lt3 = medium clayey loam; Ls4 = strong sandy loam; Sl3 = medium loamy sand; Sl4 = strong loamy sand.

```{r tab.id = "substrate", ft.align = "left", tab.cap = "Substrates"}
read_csv(here::here("data", "raw", "data_raw_table_3_substrate.csv"),
         show_col_types = FALSE) %>%
  select(-starts_with("lfl")) %>%
  slice(-1) %>%
  flextable() %>%
  bold(part = "header") %>%
  add_header_row(
    top = FALSE,
    values = read_csv(
      here::here("data", "raw", "data_raw_table_3_substrate.csv"),
                 show_col_types = FALSE) %>%
      select(-starts_with("lfl")) %>%
      slice(1)
    ) %>%
  border_inner_h(part = "header", border = fp_border(width = 1)) %>%
  align(align = "center", part = "all", j = -1) %>%
  compose(i = 1, j = 10, part = "header", value = as_paragraph("P", as_sub("2"), "O", as_sub("5"))) %>%
  compose(i = 1, j = 11, part = "header", value = as_paragraph("K", as_sub("2"), "O")) %>%
  compose(i = 1, j = 12, part = "header", value = as_paragraph("Mg", as_sup("2+"))) %>%
  compose(i = 1, j = 14, part = "header", value = as_paragraph("CaCO", as_sub("3"))) %>%
  compose(i = 2, j = 10, part = "header", value = as_paragraph("mg 100 g", as_sup("−1"))) %>%
  compose(i = 2, j = 11, part = "header", value = as_paragraph("mg 100 g", as_sup("−1"))) %>%
  compose(i = 2, j = 12, part = "header", value = as_paragraph("mg 100 g", as_sup("−1"))) %>%
  align(align = "left", part = "footer") %>%
  set_table_properties(layout = "autofit", width = 1)
```

@@

```{r fig.id = "nmds", fig.asp = 0.9696, fig.width = 6.49, fig.cap = "NMDS"}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_6_nmds_800dpi_16.5x16cm.tiff")
  )
```

NMDS ordination based on Sørensen dissimilarity of the species composition of 288 plots on dike grasslands over four years after sowing the seed mixtures (black squares).<!--# How many observed species --> In addition, 98 positive and negative reference plots were included (filled symbols) from older dike grasslands in the surroundings [@bauer2023] and six plots from sPlotOpen [@Sabatini.2021]. The circles show the standard error of the groups. 2D stress: 0.21.
